Algebraic Principles for Rely-Guarantee Style Concurrency Verification Tools

### 2013-11-15 Paper/paper.tex
Finished and submitted paper

We aim to develop lightweight formal methods which support
rely-guarantee style reasoning about concurrent programs at varying
levels of abstraction. At the most abstract level, we use algebras as
a framework for reasoning about the control flow of programs as well
as for deriving inference rules and refinement laws. At the lowest
level, concrete models of program stores support fine-grained
reasoning about program data flow and inteference. We have implemented
this approach is the Isabelle/HOL interactive theorem proving
environment. Isabelle not only provides us with a high degree of
confidence in the correctness of our development and tool prototype,
it also supports the construction of custom proof tactics and
procedures for program verification and refinement tasks. Our approach
differs from existing frameworks for rely-guarantee reasoning in
Isabelle~\cite{nieto_rely-guarantee_2003}, both by making the link
between concrete programs and algebras explicit, and by allowing
arbitrary nested parallelism.

For sequential programs, the applicability of algebra, and Kleene
algebra in particular, has been known for over a decade. Kleene
algebra provides operations for non-deterministic choice, sequential
composition and finite iteration, in addition to skip and abort. With
appropriate extensions, Kleene algebras support Hoare-style program
verification, and allow the derivation of program equivalences and
refinement rules~\cite{kozen_kleene_1997,hoare_concurrent_2011}. Kleene algebras have
been applied in applications including compiler
optimisation~\cite{kozen_certification_2000}, program
construction~\cite{berghammer_automated_2010}, transformation and
termination analysis~\cite{desharnais_algebraic_2011}, and static
analysis~\cite{fernandes_describing_2007}. Formalisations and tools are
available in interactive theorem provers such as Coq~\cite{pous_kleene_2013} and
Isabelle~\cite{armstrong_kleene_2013,armstrong_program_2013,armstrong_algebras_2014}.

Algebra also plays a role in the study of separation logic, for
example, as the separation algebras in~\cite{calcagno_local_2007}, and in algebraic
separation logic~\cite{dang_algebraic_2011}. More recently,
concurrent Kleene algebras have given an algebraic account of some
aspects of concurrent separation logic~\cite{hoare_concurrent_2011,hoare_locality_2011} and
the first steps towards an algebraic description of rely-guarantee
based reasoning have been undertaken~\cite{hoare_concurrent_2011}.

One main characteristic of the rely-guarantee method is
compositionality. The verification of large concurrent programs is
reduced to the independent verification of individual subprograms,
where the effect of interferences from other subprograms are captured
by \emph{rely} and \emph{guarantee} conditions, which constrain the
behaviour of subprograms. This form of compositionality is, however,
weaker than the strong notion of compositionality inherent in algebra,
where full congruence rules exist for every operator and refinement is
usually based on monotonicity.

In this paper, we investigate algebraic principles for rely-guarantee
style reasoning. First step have already been taken, for example,
in~\cite{hoare_concurrent_2011}, but we extend this approach by extracting a minimal set
of axioms for relies and guarantees sufficient to derive the standard
rely-guarantee inference rules. Furthermore, we explore how the
algebraic compositionality of this approach can be broken in just the
right way, so as to capture the ideas of abstract interference
conditions in rely-guarantee reasoning.

The verification and construction of actual concurrent programs often
requires complex semantics involving synchronisation, termination
analysis and intricate data structures. For simplicity, in this paper
we eschew such models in favor of relatively simple language and trace
based semantics. Despite the simplicity of this model, we demonstrate
and evaluate our prototypical verification tool implemented in
Isabelle by verifiying simple two examples from the literature. Beyond
that our approach provides a coherent framework from which more
complex and detailed models can be integrated in the future.

@@

In the context of sequential programs, one typically thinks of $\cdot$
as sequential composition, $+$ as nondeterministic choice, $0$ as the
abortive action and $1$ as skip. In this context it is essential that
multiplication is not commutative. Often we use $;$ in place of
$\cdot$ when discussing programs. More formally, it is well known that
(regular) languages with language union as $+$ and language product as
$\cdot$, the empty language as $0$ and the empty word language
$\{\varepsilon\}$ as $1$ form dioids. Another model is formed by
binary relations with the union of relations as $+$, the product of
relations as $\cdot$, the empty relation as $0$ and the identity
relation as $1$. A model of commutative dioids is formed by sets of
(finite) multisets or Parikh vectors with multiset addition as
multiplication.

@@

A \emph{trioid} is a structure $(S,+,\cdot,||,0,1)$ such that
$(S,+,\cdot,0,1)$ is a dioid and $(S,\cdot,||,0,1)$ a commutative
dioid. In a trioid there is no interaction between the sequential
composition $\cdot$ and the parallel composition $||$. On the one
hand, Gischer has shown that trioids are sound and complete for the
equational theory of series-parallel pomset languages~\cite{gischer_equational_1988},
which form a well studied model of true concurrency. On the other
hand, he has also obtained a completeness result with respect to a
notion of pomset subsumption for trioids with the additional
\emph{interchange axiom} $(w \| x) \cdot (y \| z) \le (w \cdot y) \|
(x \cdot z)$ and it is well known that this additional axiom also
holds for (regular) languages in which $||$ is interpreted as the
shuffle or interleaving operation~\cite{gischer_shuffle_1981}.

Formally, the \emph{shuffle} $\|$ of two finite words is defined
inductively as
\begin{equation*}
\epsilon \| s = \{s\}, \qquad s \| \epsilon = \{s\},\qquad
as \| bt = a(s \| bt) \cup b(as \| t),
\end{equation*}
which is then lifted to the shuffle product of languages $X$ and $Y$
as
\begin{equation*}
X \| Y = \{x \| y:\; x \in X \land x \in Y\}.
\end{equation*}

For logics of programs, notions of iteration or recursion are
essential. A \emph{Kleene algebra} is a dioid expaned with a star
operation which satisfies the \emph{left unfold axiom} $1 + x\cdot
x^*\le x^*$ and the \emph{left} and \emph{right induction axiom} $z+x
\cdot y \le y\Rightarrow x^*\cdot z \le y$ and $z + y\cdot x \le y
\Rightarrow z\cdot x^* \le y$. It follows that $1+x\cdot x^* = x^*$
and that the right unfold axiom $1+x^*\cdot x=x^*$ is derivable as
well. Thus iteration $x^\ast$ is modelled as the least fixpoint of the
function $\lambda y. 1+x\cdot y$, which is the same as the least
fixpoint of $\lambda y.1+ y\cdot x$. A \emph{commutative Kleene
  algebra} is a Kleene algebra in which multiplication is commutative.

@@

A \emph{bi-Kleene algebra} is a structure
$(K,+,\cdot,||,0,1,\hphantom{}^*,\hphantom{}^{(*)})$ where
$(K,+,\cdot,0,1,\hphantom{}^*)$ is a Kleene algebra and
$(K,+,||,0,1,\hphantom{}^{(*)})$ is a commutative Kleene
algebra. Bi-Kleene algebras are sound and complete with respect to the
equational theory of regular series-parallel pomset languages, and the
equational theory is again decidable~\cite{LaurenceStruth}. A
\emph{concurrent Kleene algebra} is a bi-Kleene algebra which
satisfies the interchange law~\cite{hoare_concurrent_2011}. It can be shown that
shuffle languages and regular series-parallel pomset languages with a
suitable notion of pomset subsumption form concurrent Kleene algebras.

In some contexts, it is also useful to add a meet operation $\sqcap$,
such that $(D,+,\sqcap)$ is a distributive lattice. This is
particularly needed in the context of refinement, where we typically
want to represent specifications as well as programs.

@@

It is well known that the inference rules for sequential Hoare logic
(except the assignment axiom) can be derived in expansions of Kleene
algebras. One approach is as follows~\cite{moller_algebras_2006}. Suppose a
suitable Boolean algebra $B$ of \emph{tests} has been embedded into a
Kleene algebra $K$ such that $0$ and $1$ are the minimal and maximal
element of $B$, $+$ corresponds to join and $\cdot$ to
meet. Complements $-$ are defined only on $B$. Suppose further that a
\emph{backward diamond operator} $\langle x|p$ has been defined for
each $x\in K$ and $p\in B$, which models the set of all states to
which each terminating execution of program $x$ may lead from states
$q$. Finally suppose that a \emph{forward box operator} $|x]p$ has
been defined which models the (largest) set of states from which every
terminating execution of $x$ must end up in states $q$ and that boxes
and diamonds are adjoints of the Galois connection
\begin{equation*}
  \langle x|p \le q \Leftrightarrow p \le |x]q,
\end{equation*}
for all $x\in K$ and $p,q\in B$. It is then evident from the above
explanations that validity of a Hoare triple $\vdash \{p\}x\{q\}$ can
be encoded as $\langle x|p \le q$ and the weakest liberal precodition
operator $\mathsf{wlp}(x,q)$ as $|x]p$. Hence the relationship between
the proof theory and the semantics of Hoare logic is captured by the
Galois connection $ \vdash\{p\}x\{q\} \Leftrightarrow p \le
\mathsf{wlp}(x,q)$. It has been shown that the relational semantics of
sequential while-programs can be encoded in these \emph{modal Kleene
  algebras} and that the inference rules of Hoare logic can be
derived~\cite{moller_algebras_2006}.

In the context of concurrency, this relational approach is no longer
appropriate. Here, the following approach by
Tarlecki~\cite{tarlecki_language_1985} can be used instead. One can
now encode validity of a Hoare triple as
\begin{equation*}
  \vdash \{x\}y\{z\} \Leftrightarrow x \cdot y \le z
\end{equation*}
for arbitrary elements of a Kleene algebra $K$. Nevertheless all the
rules of sequential Hoare logic except the assignment axiom can still
be derived in this setting~\cite{hoare_concurrent_2011}. Tarlecki's motivating
explanations carry over to the algebraic approach.

@@

The correspondence to a wlp-style semantics, as in model Kleene
algebra, now requires a generalisation of the Galois connection for
boxes and diamonds to multiplication and an upper adjoint in the form
of residuation.  This can be achieved in the context of \emph{action
  algebras}~\cite{pratt_action_1990}, which expand Kleene algebras by operations
of left and right residuation defined by the Galois connections
\begin{equation*}
  x\cdot y\le z\Leftrightarrow x\le z\leftarrow y,\qquad x\cdot y\le z \Leftrightarrow y\le x\rightarrow z.
\end{equation*}
These residuals, and now even the Kleene star can be axiomatised
equationally in action algebras. For a comprehensive list of the
properties of action algebras and their most important models
see~\cite{armstrong_kleene_2013}, including the language and the relational
model. In analogy to the development in model Kleene algebra we can
now stipulate $ \wlp(x,y)=y\leftarrow x$ and obtain the Galois
connection
\begin{equation*}
  \vdash \{x\}y\{z\} \Leftrightarrow x\le \wlp(y,z)
\end{equation*}
with the characteristic properties $\vdash \{\wlp(y,z)\}y\{z\}$ and $x\le
\wlp(y,z)\Rightarrow \ \vdash \{x\}y\{z\}$. Moreover, if the action algebra is
also a quantale, and infinite sums exist, it follows that $
\wlp(y,z)=\sum\{x:\ \vdash \{x\}y\{z\}\}$.  It is obvious that this definition
makes sense in all models of action algebras and
quantales. Intuitively, suppose $p$ stands for the set of all
behaviours of a system, for instance the set of all execution traces,
that end in state $p$, and likewise for $q$. Then $\{p\}x\{q\}$ states
that all executions ending in $p$ can be extended by $x$ to executions
ending in $q$. $\wlp(x,q)$ is the most general behaviour, that is the
set of all executions $p$ after which all executions of $x$ must end
in $q$.

In the context of concurrent Kleene algebra, even without the parallel
star, additional inference rules can be derived, in particular
\begin{equation*}
  \frac{\vdash\{x\}y\{z\}}{\vdash\{w||x\}y\{w||z\}}\qquad \frac{\vdash\{x_1\}y_1\{z_1\}\quad\vdash\{x_2\}y_2\{z_2\}}{\vdash\{x_1||x_2\}y_1||y_2\{z_1||z_2\}}
\end{equation*}
which correspond to the frame rule and the concurrency rule of
concurrent separation logic~\cite{} when $||$ is interpreted as
separating conjunction. In fact, the second inference rule is, in any
bi-Kleene algebra equivalent to the interchange axiom and the frame
rule is derivable from this axiom as well~\cite{hoare_concurrent_2011}. The
precise connection between concurrent separation logic, a variant of
concurrent Kleene algebra and a resource transformer model has been
elaborated in~\cite{hoare_locality_2011}.

@@

In this section we show how bi-Kleene algebras can be expanded into a
simple algebra that supports the derivation of rely-guarantee-style
inference rules. This development does \emph{not} use the interchange
law for several reasons. First, this law fails for fair parallel
composition $x\parallel_f y$ in a model with possibly infinite, or
non-terminating programs. In this model, $x \cdot y \not\leq x
\parallel_f y$ in the case whenever $x$ is non-terminating. Secondly,
it is not needed for deriving the usual laws or rely-guarantee.

A rely-guarantee algebra is a structure
$(K,I,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
distributive lattice, $(K,+,\cdot,\|,0,1)$ is a trioid and
$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. Note that we do not
consider the parallel star. $I$ is a distinguished subset of rely and
guarantee conditions or \emph{interference constraints} which satisfy
the following axioms
\begin{align}
r\|r &\le r, \label{rg1}\\
r &\le r\|r', \label{rg2}\\
r\|xy &= (r\|x)(r\|y), \label{rg3}\\
r\|x^+ &\le (r\|x)^+ \label{rg4}.
\end{align}
By convention, we use $r$ and $g$ to refer to elements of $I$,
depending on whether they are used as relies or guarantees, and
$x,y,z$ for arbitrary elements of $K$. The operators $\|$ and $\sqcap$
must be closed with respect to $I$.

Axiom (\ref{rg1}) states that interference from a constraint being run
twice in parallel is no different from just the interference from that
constraint begin run once in parallel. Axiom (\ref{rg2}) states that
interference from a single constraint is less than interference from
itself and another interference constraint. Axiom (\ref{rg3}) allows
an interference constraint to be split across sequential
programs. Axiom (\ref{rg4}) is similar to Axiom (\ref{rg3}) in intent,
except it deals with finite iteration.

Some elementary consequences of these rules are
\begin{equation*}
1 \le r,\qquad
r^\star = rr = r = r\|r,\qquad
r\|x^+ = (r\|x)^+.
\end{equation*}

\begin{thm}
  Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent.
\end{thm}
\begin{proof}
  To show this, we use Isabelle's
  \emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample
  generator to construct models which violate each particular axiom
  while satisfying all others.
\end{proof}

\begin{thm}
Axiom (\ref{rg4}) can be derived from (\ref{rg3}) by a simple
inductive argument in a quantale where $\|$ distributes over arbitrary suprema.
\end{thm}
\begin{proof}
  In a quantale $x^+$ can be defined as a sum of powers
  $x^+=\sum_{i\ge 1} x^i$ where $x^1=x$ and $x^{i+1}=x\cdot x^i$. By
  induction on w$i$ we get $r\|x^i = (r\|x)^i$, hence
  \begin{align*}
    r\|x^+ = r\|\sum_{i\ge 1} x^i = \sum_{i\ge 1} r\|x^i = \sum_{i\ge 1} (r\|x)^i = (r\|x)^+.
  \end{align*}
\end{proof}

@@

Jones quintuples can be encodeded in this setting as
\begin{align}
r, g \vdash \{p\} c \{q\} \iff p(r\|c) \le q \land c \le g. \label{quin}
\end{align}

\begin{thm}
With this encoding we can derive the standard rely-guarantee inference
rules, as shown in Figure \ref{fig:rgrules}.
\end{thm}

\begin{figure}[hbt]
\centering
\begin{prooftree}
\RightLabel{Skip}
\AxiomC{$pr \le p$}
\UnaryInfC{$r, g \vdash \{p\}1\{p\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{Weakening}
\AxiomC{$r' \le r$}
\AxiomC{$g \le g'$}
\AxiomC{$p \le p'$}
\AxiomC{$r', g' \vdash \{p'\}x\{q'\}$}
\AxiomC{$q' \le q$}
\QuinaryInfC{$r, g \vdash \{p\}x\{q\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{Sequential}
\AxiomC{$r, g \vdash \{p\}x\{q\}$}
\AxiomC{$r, g \vdash \{q\}y\{s\}$}
\BinaryInfC{$r, g \vdash \{q\}xy\{s\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{Parallel}
\AxiomC{$r_1, g_2 \vdash \{p_1\}x\{q_1\}$}
\AxiomC{$g_1 \le r_2$}
\AxiomC{$r_1, g_2 \vdash \{p_2\}y\{q_2\}$}
\AxiomC{$g_2 \le r_1$}
\QuaternaryInfC{$r_1 \sqcap r_2, g_1 \| g_2 \vdash \{p_1 \sqcap q_2\}x\|y\{q_1 \sqcap q_2\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{Choice}
\AxiomC{$r, g \vdash \{p\}x\{q\}$}
\AxiomC{$r, g \vdash \{p\}y\{q\}$}
\BinaryInfC{$r, g \vdash \{p\}x + y\{q\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{Star}
\AxiomC{$pr \le p$}
\AxiomC{$r, g \vdash \{p\}x\{p\}$}
\BinaryInfC{$r, g \vdash \{p\}x^\star\{p\}$}
\end{prooftree}
\caption{Rely-guarantee inference rules}
\label{fig:rgrules}
\end{figure}

Thus equations
(\ref{rg1}) to (\ref{rg4}), which are all necessary to derive these
rules, somehow represent a minimalistic set of axioms from which these
inference rules can be derived.

>>>>

Extensions of Hoare logics are becoming increasingly important for the
verification and development of concurrent and multiprocessor
programs. One of the most popular extensions is Jones' rely-guarantee
method~\cite{jones_development_1981}.  A main benefit of this method is
compositionality: the verification of large concurrent programs can be
reduced to the independent verification of individual subprograms. The
effect of interactions or interference between subprograms is captured
by \emph{rely} and \emph{guarantee} conditions. Rely conditions describe
the effect of the environment on an individual subprogram. Guarantee conditions,
in turn, describe the effect of an individual subprogram on the
environment. By constraining a subprogram by a rely condition, the global
effect of interactions is captured locally.

To make this method applicable to concrete program development and
verification tasks, its integration into tools is essential. To
capture the flexibility of the method, a number of features seem
desirable. First, we need to implement solid mathematical models for
fine-grained program behaviour. Second, we would like an abstract
layer at which inference rules and refinement laws can be derived
easily. Third, a high degree of proof automation is mandatory for the
analysis of concrete programs. In the context of the rely-guarantee
method, tools with these important features are currently missing.

This paper presents a novel approach for providing such a tool
integration in the interactive theorem proving environment
Isabelle/HOL. At the most abstract level, we use algebras to reason
about the control flow of programs as well as for deriving inference
rules and refinement laws. At the most concrete level, detailed models
of program stores support fine-grained reasoning about program data
flow and interference. These models are then linked with the
algebras. Isabelle allows us to implement these layers in a modular
way and relate them formally with one another. It not only provides us
with a high degree of confidence in the correctness of our
development, it also supports the construction of custom proof tactics
and procedures for program verification and refinement tasks.

For sequential programs, the applicability of algebra, and Kleene
algebra in particular, has been known for decades. Kleene algebra
provides operations for non-deterministic choice, sequential
composition and finite iteration, in addition to skip and abort. With
appropriate extensions, Kleene algebras support Hoare-style
verification of sequential programs, and allow the derivation of
program equivalences and refinement
rules~\cite{kozen_kleene_1997,hoare_concurrent_2011}. Kleene algebras
have been used in applications including compiler optimisation,
program construction, transformation and termination analysis, and
static analysis. Formalisations and tools are available in interactive
theorem provers such as Coq~\cite{pous_kleene_2013} and
Isabelle~\cite{armstrong_kleene_2013,armstrong_program_2013,armstrong_algebras_2014}. A
first step towards an algebraic description of rely-guarantee based
reasoning has recently been undertaken~\cite{hoare_concurrent_2011}.

The main contributions of this paper are as follows. First, we
investigate algebraic principles for rely-guarantee style
reasoning. Starting from~\cite{hoare_concurrent_2011} we extract a basic
minimal set of axioms for rely and guarantee conditions which suffice
to derive the standard rely-guarantee inference rules. These axioms
provide valuable insights into the conceptual and operational role of
these constraints. However, algebra is inherently compositional, so it
turns out that these axioms do not fully capture the semantics of
interference in execution traces. We therefore explore how the
compositionality of these axioms can be broken in the right way, so as to capture the
intended trace semantics.

Second, we link our rely-guarantee algebras with a simple trace based
semantics which so far is restricted to finite executions and
disregards termination and synchronisation. Despite the simplicity of
this model, we demonstrate and evaluate our prototypical verification
tool implemented in Isabelle by verifying two simple examples from
the literature. Beyond that our approach provides a coherent framework
from which more complex and detailed models can be implemented in the
future.

Third, we derive the usual inference rules of the rely-guarantee
method with the exception of assignment axioms directly from the
algebra, and obtain assignment axioms from our models. Our
formalisation in Isabelle allows us to reason seamlessly across these
layers, which includes the data flow and the control flow of concurrent
programs.

Taken together, our Isabelle implementation constitutes a tool
prototype for the verification and construction of concurrent
programs. We illustrate the tool with two simple examples from the
literature. The complete Isabelle code can be found
online\footnote{\url{www.dcs.shef.ac.uk/$\sim$alasdair/rg}}. A
previous Isabelle implementation of rely-guarantee reasoning is due to
Prensa Nieto~\cite{nieto_rely-guarantee_2003}. Our implementation
differs both by making the link between concrete programs and algebras
explicit, which increases modularity, and by allowing arbitrary nested parallelism.

@@

In the context of sequential programs, one typically thinks of $\cdot$
as sequential composition, $+$ as nondeterministic choice, $0$ as the
abortive action and $1$ as skip. In this context it is essential that
multiplication is not commutative. Often we use $;$ for sequential
composition when discussing programs. More formally, it is well known
that (regular) languages with language union as $+$, language
product as $\cdot$, the empty language as $0$ and the empty word
language $\{\varepsilon\}$ as $1$ form dioids. Another model is formed
by binary relations with the union of relations as $+$, the product of
relations as $\cdot$, the empty relation as $0$ and the identity
relation as $1$. A model of commutative dioids is formed by sets of
(finite) multisets or Parikh vectors with multiset addition as
multiplication.

@@

A \emph{trioid} is a structure $(S,+,\cdot,||,0,1)$ such that
$(S,+,\cdot,0,1)$ is a dioid and $(S,+,||,0,1)$ a commutative
dioid. In a trioid there is no interaction between the sequential
composition $\cdot$ and the parallel composition $||$. On the one
hand, Gischer has shown that trioids are sound and complete for the
equational theory of series-parallel pomset languages~\cite{gischer_equational_1988},
which form a well studied model of true concurrency. On the other
hand, he has also obtained a completeness result with respect to a
notion of pomset subsumption for trioids with the additional
\emph{interchange axiom} $(w \| x) \cdot (y \| z) \le (w \cdot y) \|
(x \cdot z)$ and it is well known that this additional axiom also
holds for (regular) languages in which $||$ is interpreted as the
shuffle or interleaving operation~\cite{gischer_shuffle_1981}.

Formally, the \emph{shuffle} $\|$ of two finite words is defined
inductively as $\epsilon \| s = \{s\}$, $s \| \epsilon = \{s\}$, and
$as \| bt = a(s \| bt) \cup b(as \| t)$, which is then lifted to the
shuffle product of languages $X$ and $Y$ as $X \| Y = \{x \| y:\; x
\in X \land x \in Y\}$.

For programming, notions of iteration are essential. A \emph{Kleene
  algebra} is a dioid expanded with a star operation which satisfies
both the \emph{left unfold axiom} $1 + x\cdot x^\star\le x^\star$ and
\emph{left} and \emph{right induction axioms} $z+x \cdot y \le
y\Rightarrow x^\star\cdot z \le y$ and $z + y\cdot x \le y \Rightarrow
z\cdot x^\star \le y$. It follows that $1+x\cdot x^\star = x^\star$
and that the right unfold axiom $1+x^\star\cdot x\le x^\star$ is
derivable as well. Thus iteration $x^\ast$ is modelled as the least
fixpoint of the function $\lambda y. 1+x\cdot y$, which is the same as
the least fixpoint of $\lambda y.1+ y\cdot x$. A \emph{commutative
  Kleene algebra} is a Kleene algebra in which multiplication is
commutative.

@@

A \emph{bi-Kleene algebra} is a structure
$(K,+,\cdot,||,0,1,\hphantom{}^\star,\hphantom{}^{(\star)})$ where
$(K,+,\cdot,0,1,\hphantom{}^\star)$ is a Kleene algebra and
$(K,+,||,0,1,\hphantom{}^{(\star)})$ is a commutative Kleene
algebra. Bi-Kleene algebras are sound and complete with respect to the
equational theory of regular series-parallel pomset languages, and the
equational theory is again decidable~\cite{laurence_completeness_2013}. A
\emph{concurrent Kleene algebra} is a bi-Kleene algebra which
satisfies the interchange law~\cite{hoare_concurrent_2011}. It can be shown that
shuffle languages and regular series-parallel pomset languages with a
suitable notion of pomset subsumption form concurrent Kleene algebras.

In some contexts, it is also useful to add a meet operation $\sqcap$ to a bi-Kleene algebra,
such that $(K,+,\sqcap)$ is a distributive lattice. This is
particularly needed in the context of refinement, where we typically
want to represent specifications as well as programs.

@@

It is well known that the inference rules of sequential Hoare logic
(except the assignment axiom) can be derived in expansions of Kleene
algebras. One approach is as
follows~\cite{moller_algebras_2006}. Suppose a suitable Boolean
algebra $B$ of \emph{tests} has been embedded into a Kleene algebra
$K$ such that $0$ and $1$ are the minimal and maximal element of $B$,
$+$ corresponds to join and $\cdot$ to meet. Complements $-$ are
defined only on $B$. Suppose further that a \emph{backward diamond
  operator} $\langle x|p$ has been defined for each $x\in K$ and $p\in
B$, which models the set of all states to which each terminating
execution of program $x$ may lead from states $q$. Finally suppose
that a \emph{forward box operator} $|x]p$ has been defined which
models the (largest) set of states from which every terminating
execution of $x$ must end in states $p$ and that boxes and diamonds
are adjoints of the Galois connection $\langle x|p \le q
\Leftrightarrow p \le |x]q$, for all $x\in K$ and $p,q\in B$. It is
then evident from the above explanations that validity of a Hoare
triple $\vdash \{p\}x\{q\}$ can be encoded as $\langle x|p \le q$ and
the weakest liberal precondition operator $\mathsf{wlp}(x,q)$ as
$|x]p$. Hence the relationship between the proof theory and the
semantics of Hoare logic is captured by the Galois connection $
\vdash\{p\}x\{q\} \Leftrightarrow p \le \mathsf{wlp}(x,q)$. It has
been shown that the relational semantics of sequential while-programs
can be encoded in these \emph{modal Kleene algebras} and that the
inference rules of Hoare logic can be
derived~\cite{moller_algebras_2006}.

In the context of concurrency, this relational approach is no longer
appropriate; the following approach by
Tarlecki~\cite{tarlecki_language_1985} can be used instead. One can
now encode validity of a Hoare triple as
\begin{equation*}
  \vdash \{x\}y\{z\} \Leftrightarrow x \cdot y \le z
\end{equation*}
for arbitrary elements of a Kleene algebra. Nevertheless all the
rules of sequential Hoare logic except the assignment axiom can still
be derived~\cite{hoare_concurrent_2011}. Tarlecki's motivating
explanations carry over to the algebraic approach.

@@

The correspondence to a wlp-style semantics, as in modal Kleene
algebra, now requires a generalisation of the Galois connection for
boxes and diamonds to multiplication and an upper adjoint in the form
of residuation.  This can be achieved in the context of \emph{action
  algebras}~\cite{pratt_action_1990}, which expand Kleene algebras by operations
of left and right residuation defined by the Galois connections
\begin{equation*}
  x\cdot y\le z\Leftrightarrow x\le z\leftarrow y,\qquad x\cdot y\le z \Leftrightarrow y\le x\rightarrow z.
\end{equation*}
These residuals, and now even the Kleene star, can be axiomatised
equationally in action algebras. For a comprehensive list of the
properties of action algebras and their most important models
see~\cite{armstrong_kleene_2013}, including the language and the relational
model. In analogy to the development in modal Kleene algebra we can
now stipulate $ \wlp(x,y)=y\leftarrow x$ and obtain the Galois
connection
\begin{equation*}
  \vdash \{x\}y\{z\} \Leftrightarrow x\le \wlp(y,z)
\end{equation*}
with $\vdash \{\wlp(y,z)\}y\{z\}$ and $x\le \wlp(y,z)\Rightarrow \
\vdash \{x\}y\{z\}$ as characteristic properties. Moreover, if the
action algebra is also a quantale, and infinite sums exist, it follows
that $ \wlp(y,z)=\sum\{x:\ \vdash \{x\}y\{z\}\}$.  It is obvious that
this definition makes sense in all models of action algebras and
quantales. Intuitively, suppose $p$ stands for the set of all
behaviours of a system, for instance the set of all execution traces,
that end in state $p$, and likewise for $q$. Then $\{p\}x\{q\}$ states
that all executions ending in $p$ can be extended by $x$ to executions
ending in $q$. $\wlp(x,q)$ is the most general behaviour, that is the
set of all executions $p$ after which all executions of $x$ must end
in $q$.

@@

We now show how bi-Kleene algebras can be expanded into a
simple algebra that supports the derivation of rely-guarantee style
inference rules. This development does \emph{not} use the interchange
law for several reasons. First, this law fails for fair parallel
composition $x\parallel_f y$ in models with possibly infinite, or
non-terminating programs. In this model, $x \cdot y \not\leq x
\parallel_f y$ whenever $x$ is non-terminating. Secondly, it is not
needed for deriving the usual rules of rely-guarantee.

A rely-guarantee algebra is a structure
$(K,I,+,\sqcap,\cdot,\|,\hphantom{}^\star,0,1)$, where $(K,+,\sqcap)$
is a distributive lattice, $(K,+,\cdot,\|,0,1)$ is a trioid and
$(K,+,\sqcap,\cdot,\|,\hphantom{}^\star,0,1)$ is a bi-Kleene algebra
where we do not consider the parallel star. $I$ is a distinguished
subset of rely and guarantee conditions or \emph{interference
  constraints} which satisfy the following axioms
\begin{align}
r\|r &\le r, \label{rg1}\\
r &\le r\|r', \label{rg2}\\
r\|(x\cdot y) &= (r\|x)\cdot(r\|y), \label{rg3}\\
r\|x^+ &\le (r\|x)^+ \label{rg4}.
\end{align}
By convention, we use $r$ and $g$ to refer to elements of $I$,
depending on whether they are used as relies or guarantees, and
$x,y,z$ for arbitrary elements of $K$. The operations $\|$ and $\sqcap$
must be closed with respect to $I$.

The general idea is to constrain a program by a rely condition by
executing the two in parallel. Axiom (\ref{rg1}) states that
interference from a constraint being run twice in parallel is no
different from just the interference from that constraint begin run
once in parallel. Axiom (\ref{rg2}) states that interference from a
single constraint is less than interference from itself and another
interference constraint. Axiom (\ref{rg3}) allows an interference
constraint to be split across sequential programs. Axiom (\ref{rg4})
is similar to Axiom (\ref{rg3}) in intent, except it deals with finite
iteration.

Some elementary consequences of these rules are
\begin{equation*}
1 \le r,\qquad
r^\star = r\cdot r = r = r\|r,\qquad
r\|x^+ = (r\|x)^+.
\end{equation*}

\begin{thm}
  Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent.
\end{thm}
\begin{proof}
  We have used Isabelle's
  \emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample
  generator to construct models which violate each particular axiom
  while satisfying all others. \qed
\end{proof}

\begin{thm}
  Axiom (\ref{rg3}) implies (\ref{rg4}) in a quantale
  where $\|$ distributes over arbitrary suprema.
\end{thm}
\begin{proof}
  In a quantale $x^+$ can be defined as a sum of powers
  $x^+=\sum_{i\ge 1} x^i$ where $x^1=x$ and $x^{i+1}=x\cdot x^i$. By
  induction on $i$ we get $r\|x^i = (r\|x)^i$, hence
  \begin{align*}
    r\|x^+ = r\|\sum_{i\ge 1} x^i = \sum_{i\ge 1} r\|x^i = \sum_{i\ge 1} (r\|x)^i = (r\|x)^+.
  \end{align*}
  \qed
\end{proof}

@@

Jones quintuples can be encoded in this setting as
\begin{align}
r, g \vdash \{p\} x \{q\} \iff p\cdot(r\|x) \le q \land x \le g. \label{quin}
\end{align}
This means that program $x$ when constrained by a rely $r$, and
executed after $p$, behaves as $q$. Moreover, all behaviours of $x$
are included in its guarantee.

\begin{thm}
  The standard rely-guarantee inference rules can be derived with the
  above encoding, as shown in Figure \ref{fig:rgrules}.
\end{thm}

\begin{figure}[hbt]
\centering
\begin{prooftree}
\RightLabel{\ Skip}
\AxiomC{$p\cdot r \le p$}
\UnaryInfC{$r, g \vdash \{p\}1\{p\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{\ Weakening}
\AxiomC{$r' \le r$}
\AxiomC{$g \le g'$}
\AxiomC{$p \le p'$}
\AxiomC{$r', g' \vdash \{p'\}x\{q'\}$}
\AxiomC{$q' \le q$}
\QuinaryInfC{$r, g \vdash \{p\}x\{q\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{\ Sequential}
\AxiomC{$r, g \vdash \{p\}x\{q\}$}
\AxiomC{$r, g \vdash \{q\}y\{s\}$}
\BinaryInfC{$r, g \vdash \{q\}x\cdot y\{s\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{\ Parallel}
\AxiomC{$r_1, g_2 \vdash \{p_1\}x\{q_1\}$}
\AxiomC{$g_1 \le r_2$}
\AxiomC{$r_1, g_2 \vdash \{p_2\}y\{q_2\}$}
\AxiomC{$g_2 \le r_1$}
\QuaternaryInfC{$r_1 \sqcap r_2, g_1 \| g_2 \vdash \{p_1 \sqcap q_2\}x\|y\{q_1 \sqcap q_2\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{\ Choice}
\AxiomC{$r, g \vdash \{p\}x\{q\}$}
\AxiomC{$r, g \vdash \{p\}y\{q\}$}
\BinaryInfC{$r, g \vdash \{p\}x + y\{q\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{\ Star}
\AxiomC{$p\cdot r \le p$}
\AxiomC{$r, g \vdash \{p\}x\{p\}$}
\BinaryInfC{$r, g \vdash \{p\}x^\star\{p\}$}
\end{prooftree}
\caption{Rely-guarantee inference rules}
\label{fig:rgrules}
\end{figure}

Thus (\ref{rg1}) to (\ref{rg4}), which are all necessary to derive these
rules, represent a minimal set of axioms from which these
inference rules can be derived.
